---
title: "Ch. 6 - Priors"
subtitle: "Chapter 6 Exercises"
---


# Chapter 6

```{r}
pacman::p_load(
  bcogsci,
  brms,
  dplyr,
  tibble,
  lme4
)
```

## Eliciting priors from oneself for a self-paced reading study: A simple example

An example: English relative clauses

```{r}
data("df_gg05_rc")
```


```{r}
df_gg05_rc <- df_gg05_rc %>%
  mutate(c_cond = if_else(condition == "objgap", 1 / 2, -1 / 2))
```


```{r}
head(df_gg05_rc)
```


```{r}
fit_gg05 <- brm(RT ~ c_cond + (1 + c_cond || subj) +
                  (1 + c_cond || item), df_gg05_rc,
                file = here::here("models", "exercises", "ch6", "fit_gg05"))
```

```{r}
(default_b <- posterior_summary(fit_gg05,
                                variable = "b_c_cond"))
```


Compare these estimates to a frequentist model:

```{r}
#| eval: false
fit_lmer <- lmer(RT ~ c_cond + (1 + c_cond || subj) +
                   (1 + c_cond || item), df_gg05_rc)
# store as RDS
saveRDS(fit_lmer, here::here("models", "exercises", "ch6", "fit_lmer.rds"))
```

```{r}
fit_lmer <- readRDS(here::here("models", "exercises", "ch6", "fit_lmer.rds"))
```



```{r}
b <- summary(fit_lmer)$coefficients["c_cond", "Estimate"]
SE <- summary(fit_lmer)$coefficients["c_cond", "Std. Error"]
## estimate of the slope and
## lower and upper bounds of the 95% CI:
(lmer_b <- c(b, b - (2 * SE), b + (2 * SE)))
```

They're quite similar to the Bayesian model we ran above. This is because we had used default priors, which are uninformative and don't influence the posterior much. So the likelihood dominates when determining the posteriors.

Let's use another uninformative prior, uniform priors:

```{r}
fit_gg05_unif <- brm(RT ~ c_cond + (1 + c_cond || subj) +
                       (1 + c_cond || item),
  prior = c(
    prior(uniform(-2000, 2000), class = Intercept,
          lb = -2000, ub = 2000),
    prior(uniform(-2000, 2000), class = b,
          lb = -2000, ub = 2000),
    prior(normal(0, 500), class = sd),
    prior(normal(0, 500), class = sigma)
  ), df_gg05_rc,
                file = here::here("models", "exercises", "ch6", "fit_gg05_unif")
  )
```

```{r}
(uniform_b <- posterior_summary(fit_gg05_unif, 
                                variable = c("b_c_cond")) )
```

```{r}
uniform_b <-
  uniform_b |> 
  as_tibble() |> 
   mutate(model = "Uniform")

default_b <- default_b |> 
  as_tibble() |> 
   mutate(model = "Default")

names <- names(default_b)
values <- c(lmer_b[1], SE, lmer_b[2], lmer_b[3], "Frequentist")

rbind(names, values) |> as_tibble()
```

### Eliciting a prior for the intercept

# Session Info

Compiled with `r R.version$version` (`r R.version$nickname`) in RStudio version 2023.12.1.402 (Ocean Storm).

```{r}
#| eval: false
#| echo: false
RStudio.Version()$version; RStudio.Version()$release_name
```

```{r}
sessionInfo()
```







