---
title: "Introduction to model comparison"
subtitle: "Chapter 14 notes"
date: "07/10/2024"
---

These notes accompany [Ch. 14 (Introduction to model comparison)](https://vasishth.github.io/bayescogsci/book/ch-comparison.html) of the current version of [Bayesian Data Analysis for Cognitive Science](https://vasishth.github.io/bayescogsci).

By comparing models we can compare predictions that would be posited by theories. We can compare how well theories' predictions (i.e., priors) match some data by fitting Bayesian models to each theory and comparing their predictive power. There are two perspectives on model prediction:

1. *prior* predictive perspective based on Bayes factor using marginal likelihoods
    + examines how well the model (prior + likelihood) explains the data
2. *posterior* predictive perspective based on cross-validation
    + assess model predictions for held-out data after seeing most of the data (i.e., train + test)

# Prior predictive vs. posterior predictive model comparison

Bayes factor's predictive power is based on its prior predictive distribution. 
Bayes factor is highly sensitive to priors (evaluates probability of observed data under prior assumptions). Has been called a "cruel realist" who makes judgements based on prior information, thereby penalising if we don't have good prior information.

Cross-validation fits a model to training data and posterior distributions of the parameters from this training model are then used to make predictions for test data, and model fit is assessed based no the test data (with this process repeated serveral times on different test data subsets). Cross-validations assess whether the model would generalise to new data, although our test data is not really new (e.g., collected from the same participants under the same conditions, etc.). Has been referred to as a "fair judge", in that it doesn't use prior information to inform its assessment, as this prior information may or may not be accurate and can be manipulated independently from the actual data or model itself.


# Some important points to consider when comparing models




# Session Info

Compiled with `r R.version$version` (`r R.version$nickname`) in RStudio version 2023.12.1.402 (Ocean Storm).

```{r}
#| eval: false
#| echo: false
RStudio.Version()$version; RStudio.Version()$release_name
```

```{r}
sessionInfo()
```


