---
title: "OpenHPI 2023 Set-up"
author: "Daniela Palleschi"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = F)
```

# Set-up

This script loads the required packages for the openHPI course ["Introduction to Bayesian Data Analysis"](https://open.hpi.de/courses/bayesian-statistics2023).

This script can be run prior to starting the book "An Introduction to Bayesian Data Analysis for Cognitive Science" by
Bruno Nicenboim, Daniel Schad, and Shravan Vasishth. The code is based on the section [Software needed](https://vasishth.github.io/bayescogsci/book/software-needed.html).

## CRAN packages

Required packages hosted on CRAN are installed below.

```{r}
# install packages hosted on CRAN
install.packages(c(#"SIN", # SIN package was removed from the CRAN repository
  "MASS", "dplyr", "tidyr", "purrr", "extraDistr", "ggplot2", "loo", "bridgesampling", "brms", "bayesplot", "tictoc", "hypr",   "papaja", "grid", "kableExtra", "gridExtra", "lme4", "cowplot", "pdftools", "cmdstanr", "rootSolve"))
```

Some required packages aren't available on CRAN. They are installed below.

## Developer packages

If you don't have the packages `devtools` and/or `remotes` already installed, run the next line(s). Otherwise, skip to the next chunk.

First, check to see if you have the packages `devtools`, `remotes`, `StanHeaders`, and `rstan` installed. If the packages are installed, a file path will be printed (e.g., `[1] "/Library/Frameworks/R.framework/Versions/4.1/Resources/library/remotes"`). If not, you'll get an error `Error in find.package("x") : there is no package called ‘x’`.

```{r}
find.package("devtools")
# run the next line if you don't have 'devtools' already installed
install.packages("devtools")
```


```{r}
find.package("remotes")
# run the next line if you don't have 'remotes' already installed
install.packages("devtools")
```

```{r}
find.package("StanHeaders")
# if 'Stanheaders' is installed, remove it to start fresh
remove.packages("StanHeaders")
```

```{r}
find.package("rstan")
# if 'rstan' is installed, remove it to start fresh
remove.packages("rstan")
```

Now we can install (or re-install) the relevant packages.

```{r}
# From https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started
# run the next line if you already have rstan installed
# remove.packages(c("StanHeaders", "rstan"))
install.packages("rstan", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))

# bcogsci packcage; contains data for the textbook exercises
devtools::install_github("bnicenboim/bcogsci")

# cmdstanr (https://mc-stan.org/cmdstanr/)
remotes::install_github("stan-dev/cmdstanr")

# install SBC (optional)
devtools::install_github("hyunjimoon/SBC")
```

# Loading packages in subsequent scripts

After you've first installed all packages, you can use the following code at the beginning of each subsequent script, which will load the required packages, or install them if they are missing from your library. This step only works after the packages not hosted on CRAN (rstan, bcogsci, cmdstanr) have been installed. To check that this will work in subsequent scripts, restart R ('Session > Restart R') and then run the following code chunk to see if it works.

```{r}

packages <- c( #"SIN", # this package was removed from the CRAN repository
               "MASS", "dplyr", "tidyr", "purrr", "extraDistr", "ggplot2", "loo", "bridgesampling", "brms", "bayesplot", "tictoc", "hypr", "bcogsci", "papaja", "grid", "kableExtra", "gridExtra", "lme4", "cowplot", "pdftools", "cmdstanr", "rootSolve", "SBC"
  )

# NB: if you haven't already installed bcogsci through devtools, it won't be loaded
## Now load or install & load all
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)

# this is also required, taken from the textbook

## Save compiled models:
rstan_options(auto_write = FALSE)
## Parallelize the chains using all the cores:
options(mc.cores = parallel::detectCores())
# To solve some conflicts between packages
select <- dplyr::select
extract <- rstan::extract
```

If the code above gives you trouble, just use the chunk below. It is taken from the textbook (as of Feb. 12, 2023).

```{r, eval = F}
library(MASS)
## be careful to load dplyr after MASS
library(dplyr)
library(tidyr)
library(purrr)
library(extraDistr)
library(ggplot2)
library(loo)
library(bridgesampling)
library(brms)
library(bayesplot)
library(tictoc)
library(hypr)
library(bcogsci)
library(lme4)
library(rstan)
library(cmdstanr) # This package is optional, see https://mc-stan.org/cmdstanr/
library(SBC) # This package is optional, see https://hyunjimoon.github.io/SBC/
library(SHELF)
library(rootSolve)

## Save compiled models:
rstan_options(auto_write = FALSE)
## Parallelize the chains using all the cores:
options(mc.cores = parallel::detectCores())
# To solve some conflicts between packages
select <- dplyr::select
extract <- rstan::extract
```

# Restore renv.lock

```{r}
#| eval: false
renv::restore()
```

## Update renv.lock

```{r}
#| eval: false

renv::snapshot()
```

